Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-01 paper.qmd
correct citation for MDS

<!-- - Connection between NLDR and MDS-->
Historically, \kD{}   representations of \pD{}   data have been computed using multidimensional scaling (MDS) [@borg2005], which includes principal components analysis (PCA) [@jolliffe2011] as a special case.  The \kD{}   representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{}   layout is constructed by minimizing a stress function that differences distances between points in \pD{}   with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

@@

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.2,
    color = '#a65628'
  )  +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

>>>>

<!-- - Connection between NLDR and MDS-->
Historically, \kD{}   representations of \pD{} data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (@borg2005 provides a comprehensive guide to MDS.) The \kD{}   representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{}   layout is constructed by minimizing a stress function that differences distances between points in \pD{}   with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

@@

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.2,
    size = 0.1,
    color = '#a65628'
  )  +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```
